---
title: "Field data collection and data compilation"
subtitle: "Analysing Ecological Data with Detection Error"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(ggplot2)
    library(corrplot)
    library(mefa4)
    library(skimr)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
    library(GGally)
    library(partykit)
})
```

# ABMI data

## Raw data

We start by loading the raw data file and taking a first look at its contents.
`read.csv()` imports the ABMI breeding birds table into `b`.
`str(b)` then prints a compact structural overview — number of rows and columns, each variable's name and type, and a preview of the first few values — which is the fastest way to confirm that columns were parsed correctly on import.
`skimr::skim(b)` follows with a richer per-column summary: missing value counts, completeness rates, and type-appropriate statistics such as min/max/mean for numerics and frequency counts for characters.

```{r}
if (!interactive()) {
    b <- read.csv("../data/A_T26A_Breeding_Birds_AllYear_AB.csv")
} else {
    b <- read.csv("data/A_T26A_Breeding_Birds_AllYear_AB.csv")
}

str(b)

skimr::skim(b)
```

A tidy workflow would involve the following steps:

- set the survey metadata table (site, year, coordinates, etc.)
- join the predictor variables (climate, land cover, human footprint) using dplyr's `left_join()` by site and year or using `match()` to align rows.
- join the response variables (bird counts), 1 species at a time

Alternatively, we can pivot the bird counts to a wide format, where each species is a column, and then join this data to the the predictors. Here is an example of how to do this with the `Xtab` function from the `mefa4` package, which creates a cross-tabulation of counts by site and species. The resulting object `y` will have sites as rows and species as columns, with the counts of each species at each site.

```{r}
y <- mefa4::Xtab(~ ABMI.Site + Common.Name, b)

head(y)
```

The object `y` is a so-called sparse matrix, which is a memory-efficient way to store data with many zeros (as is common in ecological count data). Use `as.matrix()` to convert it to a regular matrix if needed, e.g. before joining to the data frame with the other site variables.

## Cleaned data

In the next chunk, we switch from the raw table to a cleaned, analysis-ready data set.

First, we read the `ab_birds_site_year` object from disk. This object stores the bird counts
and the site-level variables in a format that is easier to work with than the original CSV.
It also contains the data in a way that is more suitable for modeling, with separate matrices for counts and predictors. The `d` object is a list that contains at least two components: `counts`, which is a matrix of bird counts for each species, and `vars`, which is a data frame of site-year predictors.

Next, we choose one focal species by setting `spp` to `"yellowrumped_warbler"`.
This lets us pull out a single response variable from the full community data set.

Finally, we create a new data frame called `x`. The column `y` contains the counts for the
selected species, and the remaining columns come from `d$vars`, which holds the explanatory
variables for each site-year. This gives us one table with the response and predictors side by
side, ready for plotting or modeling.

```{r}
if (!interactive()) {
    d <- readRDS("../data/ab_birds_site_year.rds")
} else {
    d <- readRDS("data/ab_birds_site_year.rds")
}

spp <- "yellowrumped_warbler"

x <- data.frame(y = d$counts[, spp], d$vars)

x$yday <- as.POSIXlt(x$fielddate)$yday / 365
x$hr <- x$fieldtime / 24
```

Variables:

- Metadata:
  - `site`, `year`, `site_year`: site, year, unique ID for site/year (879 sites, 7911 points, 2003--2014),
  - `latitude`, `longitude`: public coordinates (decimal degrees)
  - `nrname`, `nsrname`: natural regions/subregions
- Bioclimatic variables
  - `ahm`: annual heat-moisture index (scaled temp/precip),
  - `pet`: potential evapotranspiration,
  - `ffp`: frost free period
  - `map`: mean annual precipitaton,
  - `mat`: mean annual temperature,
  - `mcmt`: mean cold month (January) temperature,
  - `mwmt`: mean warm month (July) temperature.
- Land cover: natural vegetation (150 m radius circular buffer at points, averaged at sites, given as proportion [0--1] of total 7.1 ha buffer area)
  - `pforest`: proportion of forest (coniferous, mixedwood, deciduous, Treed fens and bogs, includes cutblocks),
  - `pconif`: proportion of coniferous forest (upland and lowland, includes cutblocks),
  - `pwet`: proportion of lowland (wet) areas (treed and non-treed),
  - `poldfor`: proportion of mature and old-growth (>80 yrs) forest.
- Land cover: human footprint (0-1 as well as natural vegetation)
  - `cult`: cultivation,
  - `urbind`: urban/industrial,
  - `softlin`: 'soft' (vegetated) linear features (seismic lines, pipelines, power lines, road verges),
  - `hardlin`: 'hard' (non-vegetated) linear features (road and rails),
  - `cc`: forestry cutblocks,
  - `alien`: 'alienating' disturbances (soil disturbed, non regenerating, `cult + urbind + hardlin`),
  - `succ`: 'successional' disturbances (soil undisturbed, regenerating, `cc + softlin`),
  - `thf`: total human footprint (`alien + succ`).

## Data exploration

Overview of the data frame:

```{r}
head(x) # try also tail(x)

str(x)

summary(x)

skimr::skim(x)

round(cor(x[, sapply(x, is.numeric)]), 2)

corrplot::corrplot(cor(x[, sapply(x, is.numeric)]), "ellipse")

heatmap(cor(x[, sapply(x, is.numeric)]),
    symm = TRUE,
    col = hcl.colors(100, "RdYlBu", rev = TRUE)
)
```

Explore the response variable:

```{r}
table(x$y)

plot(table(x$y))
```

Explore the predictor variables:

```{r}
hist(x$ahm)
hist(x$pforest)
```

Bivariate relationships:

```{r}
# where to begin
cor(x[, sapply(x, is.numeric)])[, "y"] |> sort()

x |>
    ggplot(aes(x = pforest, y = y)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", se = TRUE)

x |>
    ggplot(aes(x = pconif, y = y)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", se = TRUE)

x |>
    ggplot(aes(x = as.factor(round(wind)), y = y)) +
    geom_boxplot()
```

Trivariate relationships:

```{r}
x |>
    ggplot(aes(x = pforest, y = ahm, col = y)) +
    geom_point()

x |>
    ggplot(aes(x = longitude, y = latitude)) +
    geom_point(aes(color = y > 0)) +
    coord_quickmap()

x |>
    ggplot(aes(x = yday, y = hr, col = y)) +
    geom_point()
# need to filter some extreme values for better visualization
x <- x |>
    dplyr::filter(
        hr > 0.1,
        hr < 0.6,
        yday > 0.37,
        yday < 0.5
    )
x |>
    ggplot(aes(x = yday, y = hr, col = y)) +
    geom_point()
```

Multivariate relationships:

```{r}
GGally::ggpairs(
    x,
    columns = c("y", "ahm", "pforest", "pconif", "thf", "wind", "yday", "hr"),
    diag = list(continuous = GGally::wrap("barDiag"))
)
```

# JOSM data

## Covariates

Variables that co-vary with the response variable.
Also called as independent variables, predictors.

Let's continue with the JOSM data set:

```{r}
x <- detect::josm$surveys |>
    dplyr::select(
        Longitude,
        Latitude,
        WindStart,
        TSSR,
        DAY,
        Open,
        Water,
        Decid,
        OpenWet,
        Conif,
        ConifWet,
        Agr,
        UrbInd,
        SoftLin,
        Roads
    )
```

STOP AND EXPLAIN EACH VARIABLE!

## Variable types

What type of variables we have? You can use the `str()` function to
reveal the structure of R objects:

```{r}
str(x)
```

We see that these variables are all continuous.

However, `WindStart` has very few distinct values, so we could treat it
as ordinal (ordered factor):

```{r}
table(x$WindStart)
x$WindOrd <- as.ordered(x$WindStart)
str(x$WindOrd)
levels(x$WindOrd)
```

Sometimes variables are binary. In R these can be logical (`TRUE`/`FALSE`)
or coded as 0/1. Often we make such variables by discretizing other
continuous or ordinal variables. E.g. We can create a binary wind variable:

```{r}
x$Wind01 <- ifelse(x$WindStart > 0, 1, 0)
table(x$WindStart, x$Wind01)
```

Some categorical variables depend on some kind of classification,
for example we can cut a continuous variable into bins:

```{r}
x$DecidCut <- cut(x$Decid, seq(0, 1, 0.2), include.lowest = TRUE)
table(x$DecidCut)
boxplot(Decid ~ DecidCut, x)
```

We can also inspect the land cover proportions that add up to 1 for each row.

```{r}
## define column names
cn <- c(
    "Open", "Water", "Agr", "UrbInd", "SoftLin", "Roads", "Decid",
    "OpenWet", "Conif", "ConifWet"
)
## these sum to 1
summary(rowSums(x[, cn]))
```

The `find_max()` function finds the maximum value in each row, the output 
contains the value and the column where it was found, we can turn that into
the dominant land cover type encoded in `HAB`:

```{r}
h <- mefa4::find_max(x[, cn])
head(h)
hist(h$value)
table(h$index)
x$HAB <- droplevels(h$index) # drop empty levels
x$DEC <- ifelse(x$HAB == "Decid", 1, 0)
```

Other types of categorical variables are truly discrete, like observer,
where there are no underlying continuous data:

```{r}
table(detect::josm$surveys$ObserverID)
```

When the number of categories increase and approach the sample size,
we can consider treating these variables as random effects. E.g. the 
`SurveyArea` variable that has 271 levels.

## Data exploration

We should inspect each variable that we want to use as a covariate.
Here are some of the most important functions:

```{r}
summary(x$Decid) # check mean, range, missing values
hist(x$Decid) # check skew and outliers
```

A nice way of getting all of the above nicely formatted is to use the skimr package:

```{r}
skimr::skim(x)
```

To explore relationships between variables, make scatter and box plots:

```{r}
x |> ggplot(aes(x = Decid, y = Conif)) +
    geom_point()
x |> ggplot(aes(x = Longitude, y = Latitude)) +
    geom_point()
x |> ggplot(aes(x = DAY, y = TSSR)) +
    geom_point()
x |> ggplot(aes(x = HAB, y = Decid)) +
    geom_boxplot()
```

We can present 3 variables as color scatter or bubble plots:

```{r}
x |> ggplot(aes(x = Decid, y = ConifWet, col = HAB)) +
    geom_point()
```

Multivariate exploration include checking correlations:

```{r}
round(cor(x[, cn]), 3)

corrplot::corrplot(cor(x[, cn]), "ellipse")

heatmap(as.matrix(x[, cn]))
```

## Variable transformations

We have seen examples of these:

- indicator variables (0/1)
- discretization (cut)

Other transformations include:

- sqrt: to tame outliers (not suitable for negative values)
- log: we'll see many use cases later
- polynomials: nonlinear terms (`x^2`, `x^3`, etc.)
- centering: keeps the distribution but shifts the mean
- scaling: keeps the distribution but shifts the range (often used with centering)

## Compound variables

We can reduce correlation by combining variables together when those are additive:

```{r}
x$FOR <- x$Decid + x$Conif + x$ConifWet
x$HF <- x$Agr + x$UrbInd + x$Roads + x$SoftLin
x$WET <- x$OpenWet + x$ConifWet + x$Water
```

We can reduce colinearity by calculating variables relative to each other,
like proportions or ratios (watch out for division by 0):

```{r}
x$pDecid <- ifelse(x$FOR > 0, x$Decid / x$FOR, 0)
cor(x[, c("FOR", "pDecid")])
```

We can also merge categories, e.g. based on their similarity:

```{r}
heatmap(cor(x[, cn]))
```

## Missing data

Let's check for missing values:

```{r}
data.frame(missing = colSums(is.na(x))) |> filter(missing > 0)
```

There are only 2 rows with missing values, we could use
`na.omit(x)` to drop these, or we can impute the values:

- pick a value randomly
- use the mean
- use the most common value (mode)
- Use other variables to predict the possible value

After inspecting the full data set, it turns out that site FT21204W
experiences wind according to `WindEnd`:

```{r}
sites <- rownames(x)[is.na(x$WindStart)]
detect::josm$surveys[detect::josm$surveys$SiteID %in% c("FT07424", "FT21204W"), ]
```

Let's set the value to 3 for site FT21204W, and the mode (0) for site FT07424:

```{r}
x[sites, "WindStart"] <- c(0, 3)
x[sites, "WindOrd"] <- c("0", "3")
x[sites, "Wind01"] <- c(0, 1)
```

Any more `NA`s left?

```{r}
any(is.na(x))
```

## Response variable

We have manipulated the covariates and the species counts. Let's put them together
in the same data frame.

```{r}
spp <- "OVEN" # change here if you want to use another species
detect::josm$species[spp, ]

y <- mefa4::Xtab(~ SiteID + SpeciesID, detect::josm$counts)[, spp, drop = FALSE]
x$Count <- y[rownames(x), ]
```

Let's inspect the response variable:

```{r}
table(x$Count)
```

_Question_: How would we model the abundance of this species?
