---
title: "Overview of regression techniques"
subtitle: "Analysing Ecological Data with Detection Error"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(pscl)
    library(lme4)
    library(mgcv)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
    library(visreg)
})

if (!interactive()) {
    d <- readRDS("../data/ab_birds_site_year.rds")
} else {
    d <- readRDS("data/ab_birds_site_year.rds")
}

spp <- "yellowrumped_warbler"

x <- data.frame(y = d$counts[, spp], d$vars)

x$yday <- as.POSIXlt(x$fielddate)$yday / 365
x$hr <- x$fieldtime / 24
x <- x |>
    dplyr::filter(
        hr > 0.1,
        hr < 0.6,
        yday > 0.37,
        yday < 0.5
    )
```

# Regression analysis

Almost everything in statistics can be formulated as a regression
analysis problem.

We are making the following assumptions about the data:

- the data are representative sample from the population
- the data are independent of each other and are identically distributed

These are count data and hence a natural model for them is a
discrete distribution such as Poisson or Negative Binomial.
However, for the sake of initial analysis, we will use a Normal
distribution by using the log-transformed data (after adding 0.5 to
the 0’s). The histogram indicates that it might be approximately
symmetric (except for the peak indicating the original zeros).

```{r}
x$log_y <- log(x$y + 0.5)

hist(x$log_y)
```

Let us assume that the data are approximately bell shaped and a
Gaussian distribution fits it well. In mathematical notation, we write
this as:

$Y_i \sim Normal(\mu, \sigma^2)$

where $Y_i$ is the random variable representing the log-transformed count at site $i$, $\mu$ is the mean of the distribution, and $\sigma^2$ is the variance. The observations $y_i$ are realizations of these random variables.

```{r}
m0 <- lm(log_y ~ 1, data = x)

summary(m0)
summary(m0)$sigma # residual standard deviation
summary(m0)$r.squared # R^2

coef(m0)
mean(x$log_y)
mean(fitted(m0))
```

I The best point predictor for the population abundance at a new
location is the mean of the population, namely, μ.

I We use the sample mean to predict the population abundance at a
new location. In our data example, this turns out to be
approximately exp( ˆμ) = 5.37

Is that a sufficient answer for a manager? We suspect not

We can differentiate between the following types of uncertainties:

1. If we had a slightly different sample, the sample mean would be
different. This is the estimation uncertainty. How different would
these sample means be? This is answered by the **confidence interval**
2. Even if we knew the population mean, we still have uncertainty
about the specific observation at a new location. This is because of
the intrinsic uncertainty characterized by $\sigma$
3. Uncertainty in the prediction of an observation at a new location is a
combination of the estimation uncertainty and the intrinsic
uncertainty, the **prediction interval**
1. There is another uncertainty, that of not knowing that the
true distribution is a Normal distribution, this is model uncertainty

The estimation uncertainty can be reduced
by using good sampling designs and by increasing the sample size.

_Question_: What constitutes as a good sampling design?

Intrinsic uncertainty is unaffected by sampling design and
sample size. This can be reduced only by using an appropriate model
and useful auxiliary information. Model selection and choice of
appropriate covariates is critical to reduce the intrinsic uncertainty.

Let us see how we can use auxiliary information in the above problem. As
biologists, we know that there are a number of different habitat
characteristics that affect the population abundance.

1. Can we use these habitat characteristics to reduce the intrinsic
variability?
1. How do we decide which habitat characteristics are the most
important?

There are a few exploratory techniques that we find useful in making
these decisions. We will describe them and illustrate their use in the data
analysis.

Classification and Regression Trees (CART): This technique is very
useful in ranking different covariates in terms of their use in
prediction. The actual implementation of the technique involves a
number of decisions that are made automatically (black-box) in the
algorithm. These may or may not be appropriate for every data set.
We use this technique for screening the covariates.

```{r}
ct <- x |>
    dplyr::select(
        -y,
        -site, -year, -site_year,
        -fielddate, -fieldtime,
        -nrname, -nsrname
    ) |>
    partykit::ctree(
        log_y ~ .,
        data = _
    )

ct

plot(ct)
```

Generalized Additive Models (GAM): This technique is useful in
determining if there are any non-linearities in the relationship
between the covariates and the response variable.

```{r}
gm <- mgcv::gam(
    log_y ~ s(pconif) + s(ahm) + s(thf),
    data = x
)

plot(gm, pages = 1)
mgcv::vis.gam(gm, c("pconif", "ahm"), plot.type = "contour")
mgcv::vis.gam(gm, c("pconif", "ahm"), plot.type = "persp")
```

We now analyze the data using the most important covariate
according to the CART analysis. We will see that using auxiliary
information reduces the prediction error.

1. We keep on adding the covariates until the prediction error is
sufficiently small.
1. Adding the covariates decreases the intrinsic variability but increases
the estimation uncertainty. The model selection approach has to
balance between these two trade-offs. A commonly used method for
model selection is based on the Akaike Information Criterion (AIC)
and its variants.
1. This can be done in a forward or backward fashion. The two answers
do not always match (in fact, they seldom match).

## Main effects

```{r}
m1 <- lm(log_y ~ pconif, data = x)
m2 <- lm(log_y ~ pconif + ahm, data = x)
m3 <- lm(log_y ~ pconif + ahm + thf, data = x)

anova(m0, m1, m2, m3)
AIC(m0, m1, m2, m3)
MuMIn::model.sel(m0, m1, m2, m3)
```

## Nonlinear effects

```{r}
m4 <- lm(log_y ~ pconif + I(pconif^2), data = x)
m5 <- lm(log_y ~ pconif + I(pconif^2) + ahm, data = x)
m6 <- lm(log_y ~ pconif + I(pconif^2) + ahm + thf, data = x)

MuMIn::model.sel(m0, m1, m2, m3, m4, m5, m6)
```

## Categorical variables

```{r}
# reference (1st) level is Boreal
m7 <- lm(log_y ~ nrname, data = x)
summary(m7)

MuMIn::model.sel(m0, m1, m2, m3, m4, m5, m6, m7)
```

## Interactions

```{r}
m8 <- lm(log_y ~ pconif * ahm, data = x)
summary(m8)

MuMIn::model.sel(m0, m1, m2, m3, m4, m5, m6, m7, m8)
```

Notice that as model complexity increases, the estimation uncertainty increases as well. We have to balance between the two. AIC and its variants are commonly used for this purpose, but they are not the only method. We can also use cross-validation to estimate the prediction error of the models and select the one with the lowest error.

```{r}
summary(m0)
confint(m0)

summary(m6)
confint(m6)

# marginal effects
data.frame(x, fit = fitted(m6)) |>
    ggplot(aes(x = pconif, y = fit)) +
    geom_jitter() +
    geom_smooth()
data.frame(x, fit = fitted(m6)) |>
    ggplot(aes(x = thf, y = fit)) +
    geom_jitter() +
    geom_smooth()
data.frame(x, fit = fitted(m6)) |>
    ggplot(aes(x = ahm, y = fit)) +
    geom_jitter() +
    geom_smooth()

# partial effects
visreg::visreg(m1, "pconif")
visreg::visreg(m6, "pconif")
visreg::visreg2d(m6, "pconif", "ahm")
visreg::visreg2d(m8, "pconif", "ahm")
visreg::visreg(m7, "nrname")
```

Exercise: Play with other covariates and their interactions. Try to find a better model using AIC/AICc. Make sure that effects are sensible using `summary()` and visualizations.

# Generalized Linear Models

1. Our goal in regression is to model the intrinsic variability
appropriately and to reduce it using the auxiliary information.
2. Intrinsic variability in a response variable may not be appropriately
modelled by a Normal distribution.
3. If the response variable is discrete, such as counts, we should use
appropriate statistical model such as a discrete distribution to model
intrinsic variability.
4. If the response variable is categorical, such as presence/absence of a
species, we should use an appropriate distribution such as a
Bernoulli distribution.

- We assume that the mean of the distribution can be modelled as a
function of the covariates.
- We do NOT transform the response variable. It remains on its
original scale.
- The variance of the response variable is not independent of the
mean. In fact, in many cases, it varies with the mean. The
assumption of constant variance in the linear regression is not
necessary.

We will start by looking at the abundance data that we analyzed in
the morning session. However, this time we will not transform the
response variable.

The response variable is counts taking values in (0,1,2,...). A
natural model to model the intrinsic variation is the Poisson
distribution. One can also use Negative Binomial distribution or
some such distribution. Let us see if we can use this simple
distribution to predict abundance at a new location and obtain its
prediction uncertainty.

If this uncertainty is small, we will not bother doing any further
analysis. If this uncertainty is large, we will use auxiliary information
to reduce it. The steps we follow are identical to the ones we did in
the morning session.

Mean of the Poisson distribution is always positive. Hence we model
the relationship between the mean of the Poisson distribution and
the auxiliary information using a function that is always positive.
Any such function will suffice. The most commonly used function is
the log-linear function.

We use AIC to find the best fitting model that balances the
estimation uncertainty and intrinsic variability.

We observe that the parameter estimates obtained by the GLM are
similar but not the same as the ones we obtained by using LM on the
log-transformed data. The differences between two methods are (1) No
issues with 0 counts, (2) Prediction error is more appropriate on the
discrete scale than the artificial continuous scale.

```{r}
mP0 <- glm(y ~ 1, data = x, family = poisson)
summary(mP0)
coef(mP0)
confint(mP0)
```


```{r}
ct2 <- x |>
    dplyr::select(
        -log_y,
        -site, -year, -site_year,
        -fielddate, -fieldtime,
        -nrname, -nsrname
    ) |>
    partykit::ctree(
        y ~ .,
        data = _
    )

ct2

plot(ct2)
```

```{r}
mP1 <- glm(y ~ pconif, data = x, family = poisson)
summary(mP1)
coef(mP1)
confint(mP1)
anova(mP1, test = "Chi")

mP2 <- glm(y ~ pconif + I(pconif^2), data = x, family = poisson)
summary(mP2)
coef(mP2)
confint(mP2)
anova(mP2, test = "Chi")

mP3 <- glm(y ~ pconif + I(pconif^2) + ahm, data = x, family = poisson)
summary(mP3)
coef(mP3)
confint(mP3)
anova(mP3, test = "Chi")

mP4 <- glm(y ~ pconif + I(pconif^2) + ahm + thf, data = x, family = poisson)
summary(mP4)
coef(mP4)
confint(mP4)
anova(mP4, test = "Chi")

MuMIn::model.sel(mP0, mP1, mP2, mP3, mP4)

# marginal effects
data.frame(x, fit = fitted(mP4)) |>
    ggplot(aes(x = pconif, y = fit)) +
    geom_jitter() +
    geom_smooth()
data.frame(x, fit = fitted(mP4)) |>
    ggplot(aes(x = ahm, y = fit)) +
    geom_jitter() +
    geom_smooth()
data.frame(x, fit = fitted(mP4)) |>
    ggplot(aes(x = thf, y = fit)) +
    geom_jitter() +
    geom_smooth()

# partial effects
visreg::visreg(mP1, "pconif")
visreg::visreg(mP4, "pconif")
visreg::visreg2d(mP4, "pconif", "ahm")
```


Now let us use the same data but now instead of observing the actual
counts, suppose we only noted 0’s and 1’s where 1 indicates that the
count was anything other than 0. Thus, we have turned the count data
into a binary data. These kinds of data are not quite common in
practice. How should we analyze these data?

We follow exactly the same steps as before.
1. We fit the model without any auxiliary information. How good are
the predictions?
1. Prediction accuracy for binary data is tricky to define. A common
measure of the prediction accuracy is based on how often we predict
1’s as 1’s (Sensitivity) and 0’s and 0’s (Specificity).
1. We use the Receiver Operating Characteristic (ROC) curve to show
the predictive ability. The area under the ROC curve is a measure of
predictive ability of a classification model. We want this area to be
close to 1.
Are the results comparable to LM and Poisson regression? Notice that
we lose substantial amount of information when we go from count data
to binary data.

```{r}
x$y01 <- ifelse(x$y > 0, 1, 0)
mB0 <- glm(y01 ~ 1, data = x, family = binomial)
summary(mB0)
coef(mB0)
confint(mB0)
```


```{r}
ct3 <- x |>
    dplyr::select(
        -log_y, -y,
        -site, -year, -site_year,
        -fielddate, -fieldtime,
        -nrname, -nsrname
    ) |>
    partykit::ctree(
        y01 ~ .,
        data = _
    )

ct3

plot(ct3)
```


```{r}
mB1 <- glm(y01 ~ pconif, data = x, family = binomial)
summary(mB1)
coef(mB1)
confint(mB1)
anova(mB1, test = "Chi")

mB2 <- glm(y01 ~ pconif + I(pconif^2), data = x, family = binomial)
summary(mB2)
coef(mB2)
confint(mB2)
anova(mB2, test = "Chi")

mB3 <- glm(y01 ~ pconif + I(pconif^2) + ahm, data = x, family = binomial)
summary(mB3)
coef(mB3)
confint(mB3)
anova(mB3, test = "Chi")

mB4 <- glm(y01 ~ pconif + I(pconif^2) + ahm + thf, data = x, family = binomial)
summary(mB4)
coef(mB4)
confint(mB4)
anova(mB4, test = "Chi")

MuMIn::model.sel(mB0, mB1, mB2, mB3, mB4)

# marginal effects
data.frame(x, fit = fitted(mB4)) |>
    ggplot(aes(x = pconif, y = fit)) +
    geom_jitter() +
    geom_smooth()
data.frame(x, fit = fitted(mB4)) |>
    ggplot(aes(x = ahm, y = fit)) +
    geom_jitter() +
    geom_smooth()
data.frame(x, fit = fitted(mB4)) |>
    ggplot(aes(x = thf, y = fit)) +
    geom_jitter() +
    geom_smooth()

# partial effects
visreg::visreg(mB1, "pconif")
visreg::visreg(mB4, "pconif")
visreg::visreg2d(mB4, "pconif", "ahm")
```

Notice the following:
1. The logic behind LM and GLM is nearly identical. Our goal is to
reduce the prediction uncertainty using auxiliary information.
2. The main point is that we have to use appropriate statistical model
to quantify the intrinsic variability in the response variable.
3. In both LM and GLM, we model the mean as a function of the
covariates. This function has to conform to the natural limits
imposed by the statistical model. The mean has to be positive for
positive random variables; between 0 and 1 for a binary random
variable etc.
4. Concept of residuals is extremely tricky when the range of the
statistical model is constrained.

## Count regression models

We'll use the JOSM data set.

Assemble the JOSM data:

```{r}
x <- detect::josm$surveys |>
    select(
        Longitude,
        Latitude,
        WindStart,
        TSSR,
        DAY,
        Open,
        Water,
        Decid,
        OpenWet,
        Conif,
        ConifWet,
        Agr,
        UrbInd,
        SoftLin,
        Roads
    )
x[is.na(x$WindStart), "WindStart"] <- c(0, 3)

x$FOR <- x$Decid + x$Conif + x$ConifWet
x$HF <- x$Agr + x$UrbInd + x$Roads + x$SoftLin
x$WET <- x$OpenWet + x$ConifWet + x$Water
cn <- c(
    "Open", "Water", "Agr", "UrbInd", "SoftLin", "Roads", "Decid",
    "OpenWet", "Conif", "ConifWet"
)
h <- mefa4::find_max(x[, cn])
x$HAB <- droplevels(h$index) # drop empty levels
x$DEC <- ifelse(x$HAB == "Decid", 1, 0)

spp <- "OVEN" # change here if you want to use another species

y <- mefa4::Xtab(~ SiteID + SpeciesID, detect::josm$counts)[, spp, drop = FALSE]
x$Count <- y[rownames(x), ]
```

Variables are similar to the ABMI example.

Let's inspect the response variable:

```{r}
table(x$Count)

plot(table(x$Count))
```

The distribution looks somewhat 0 inflated. What are the possible reasons for that?

- conditions lead to the absence of the species
- detected counts are lower than the actual counts

The types of models used most often to model count data and the functions used to fit them:

- Poisson: `stats::glm()`
- Negative Binomial (higher or lower variance than Poisson): `MASS::glm.nb()`
- Zero-inflated Poisson (ZIP): `pscl::zeroinfl()`
- Zero-inflated Negative Binomial (ZINB): : `pscl::zeroinfl()`

Other approaches include additive models (`mgcv::gam()`) and
tree based methods (see the gbm and xgboost packages).

Let us start with the Poisson model.

### Poisson null model

The null model states that the expected values of the count at all locations 
are identical: $E[Y_i]=\lambda$ ($i=1,...,n$), where $Y_i$ is a random variable 
that follows a Poisson distribution with mean $\lambda$: 
$(Y_i \mid \lambda) \sim Poisson(\lambda)$. The observation ($y_i$) is a 
realization of the random variables $Y$ at site $i$, these observations are 
independent and identically distributed (i.i.d.), and we have $n$ observations in total.

Saying the the distribution is Poisson is an assumption in itself. 
For example we assume that the variance equals the mean ($V(\mu)=\mu$).

```{r}
mP0 <- glm(Count ~ 1, data = x, family = poisson)
```

The `family=poisson` specification implicitly assumes that we use a logarithmic 
link functions, that is to say that $log(\lambda) = \beta_0$, or equivalently: 
$\lambda = e^{\beta_0}$. The mean of the observations equal the mean of the 
fitted values, as expected:

```{r}
mean(mP0$y)
mean(fitted(mP0))
exp(coef(mP0))
```

The logarithmic function is called the link function, its inverse, the 
exponential function is called the inverse link function. 
The model family has these conveniently stored for us:

```{r}
mP0$family
mP0$family$linkfun
mP0$family$linkinv
```

Inspect the summary

```{r}
summary(mP0)
```

Notice that the residual deviance much higher than residual degrees of freedom.
This indicates that our parametric model (Poisson error distribution, 
constant expected value) is not quite right. See if we can improve that somehow
and explain more of the variation.

We can pick an error distribution that would fit the residuals around the 
constant expected value better (i.e. using random effects). But this way we 
would not learn about what is driving the variation in the counts. 
We would also have a really hard time predicting abundance of the species for 
unsurveyed locations. We would be right on average, but we wouldn't be able to 
tell how abundance varies along e.g. a disturbance gradient or with tree cover.

An alternative approach would be to find predictors that could explain the variation.

## Main effects

We fit a parametric (Poisson) linear model using `Decid` as a predictor:

```{r}
mP1 <- glm(Count ~ Decid, data = x, family = poisson)
mean(mP1$y)
mean(fitted(mP1))
coef(mP1)
```

Same as before, the mean of the observations equal the mean of the fitted values. 
But instead of only the intercept, now we have 2 coefficients estimated. 
Our linear predictor thus looks like:
$log(\lambda_i) = \beta_0 + \beta_1 x_{1i}$. 
This means that expected abundance is
$e^{\beta_0}$ where `Decid`=0, $e^{\beta_0}e^{\beta_1}$
where `Decid`=1, and $e^{\beta_0+\beta_1 x_{1}}$ in between.

The relationship can be visualized by plotting the fitted values against the 
predictor, or using the coefficients to make predictions using our formula:

```{r}
## make a sequence between 0 and 1
dec <- seq(from = 0, to = 1, by = 0.01)
## predict lambda
lam <- exp(coef(mP1)[1] + coef(mP1)[2] * dec)

plot(fitted(mP1) ~ Decid, x, pch = 19, col = "grey") # fitted
lines(lam ~ dec, col = 2) # our predicted
rug(x$Decid) # observed x values
```

The model summary tells us that residuals are not quite right 
The residual deviance is still higher than residual degrees of freedom 
(these should be close if the Poisson assumption holds, but it is much better 
than what we saw for the null model).

We also learned that the `Decid` effect is significant 
(meaning that the effect size is large compared to the standard error):

```{r}
summary(mP1)
```

Note: we see a significant (<0.05) $P$-value for the intercept as well. 
It is totally meaningless. That $P$-value relates to the hull hypothesis of the 
intercept ($\beta_0$) being 0. There is nothing special about that, it is like 
saying the average abundance is different from 1. 

But when $\beta_1$ is significantly different from 0, 
it means that the main effect has non-negligible effect on the mean abundance.

We can compare this model to the null (constant, intercept-only) model:

```{r}
AIC(mP0, mP1)
BIC(mP0, mP1)
MuMIn::model.sel(mP0, mP1)
```

AIC uses the negative log likelihood and the number of parameters as penalty. 
Smaller value indicate a model that is closer to the (unknowable) true model 
(caveat: this statement is true only asymptotically, i.e. it holds for large
sample sizes). For small samples, we often use 
BIC (more penalty for complex models when sample size is small), or 
AICc (as in `MuMIn::model.sel()`).

### Non-linear effects

We can use polynomial terms to capture non (log) linear effects:

```{r}
mP12 <- glm(Count ~ Decid + I(Decid^2), data = x, family = poisson)
mP13 <- glm(Count ~ Decid + I(Decid^2) + I(Decid^3), data = x, family = poisson)
mP14 <- glm(Count ~ Decid + I(Decid^2) + I(Decid^3) + I(Decid^4), data = x, family = poisson)
MuMIn::model.sel(mP0, mP1, mP12, mP13, mP14)
```

Not a surprise that the most complex model won, we had enough degrees of freedoms to spare. 

```{r}
xnew <- data.frame(Decid = seq(0, 1, 0.01))

pr <- cbind(
    predict(mP1, xnew, type = "response"),
    predict(mP12, xnew, type = "response"),
    predict(mP13, xnew, type = "response"),
    predict(mP14, xnew, type = "response")
)
matplot(xnew$Decid, pr,
    lty = 1, type = "l",
    xlab = "Decid", ylab = "E[Y]"
)
legend("topleft",
    lty = 1, col = 1:4, bty = "n",
    legend = c("Linear", "Quadratic", "Cubic", "Quartic")
)
```

### Categorical variables

Categorical variables are expanded into a _model matrix_ before parameter estimation. 
The model matrix usually contains indicator variables for each level 
(value 1 when factor value equals a particular label, 0 otherwise) except for 
the _reference category_ (check `relevel` if you want to change the reference category).

The estimate for the reference category comes from the intercept, the rest of 
the estimates are relative to the reference category. In the log-linear model 
example this means a ratio.

```{r}
head(model.matrix(~DEC, x))
mP2 <- glm(Count ~ DEC, data = x, family = poisson)
summary(mP2)
coef(mP2)
```

The estimate for a non-deciduous landscape is $e^{\beta_0}$, and it is 
$e^{\beta_0}e^{\beta_1}$ for deciduous landscapes. 
(Of course such binary classification at the landscape (1 km$^2$) level doesn't really makes sense.)

```{r regr-pois_cat1}
MuMIn::model.sel(mP1, mP2)
```

Having estimates for each land cover type improves the model,
but the  model using continuous variable is still better

```{r regr-pois_cat2}
mP3 <- glm(Count ~ HAB, data = x, family = poisson)
summary(mP3)

MuMIn::model.sel(mP1, mP2, mP3)
```

The prediction in this case would look like: 
$log(\lambda_i)=\beta_0 + \sum_{j=1}^{k-1} \beta_j x_{ji}$, where we have $k$ 
factor levels (and $k-1$ indicator variables besides the intercept).

Here is a general way of calculating fitted values or making predictions based 
on the design matrix (`X`) and the coefficients (`b`) (column ordering in `X` 
must match the elements in `b`) given a parametric log-linear model `object` 
and data frame `df` (the code won't run as is, `object` is just a placeholder 
for your GLM model object):

```{r regr-pred_general,eval=FALSE}
b <- coef(object)
X <- model.matrix(formula(object), df)
exp(X %*% b)
```

### Multiple main effects

We can add main effects by providing a new formula to a new model.
Or we can use the `update()` function. The `. ~ .` means to keep
both sides of the formula but add (`+`) or remove (`-`) terms from
the original scope.
Note: the `update` function can be used to update any function call,
not just GLM models.

```{r}
# mP4 <- glm(Count ~ Decid + ConifWet, data=x, family=poisson)
mP4 <- update(mP1, . ~ . + ConifWet)
summary(mP4)

MuMIn::model.sel(mP0, mP1, mP4)
```

Here are some functions that can automate model selection:

- `drop1`: evaluate which variable to drop to lower AIC the most
- `add1`: evaluate which variable to add to lower AIC the most
- `step`: perform forward/backward model selection using `add1`/`drop1`

### Interactions

When we consider interactions between two variables (say $x_1$ and $x_2$), 
we refer to adding another variable to the model matrix that is a product of 
the two variables ($x_{12}=x_1 x_2$):

```{r}
head(model.matrix(~ x1 * x2, data.frame(x1 = 1:4, x2 = 10:7)))
```

Let's consider interaction between our two predictors from before:

```{r regr-inter}
mP5 <- glm(Count ~ Decid * ConifWet, data = x, family = poisson)
summary(mP5)
MuMIn::model.sel(mP0, mP1, mP4, mP5)
```

The model with the interaction is best supported, but how do we make sense of 
this relationship? We can't easily visualize it in a single plot. We can either

1. fix all variables (at their mean/meadian) and see how the response is changing along a single variable: this is called a _conditional_ effect (conditional on fixing other variables), this is what `visreg::visreg()` does
2. or plot the fitted values against the predictor variable (one at a time), this is called a _marginal_ effect, and this is what `ResourceSelection::mep()` does

```{r regr-visreg2}
visreg::visreg(mP5, scale = "response", xvar = "ConifWet", by = "Decid")
```

```{r}
ggplot(x, aes(x = Decid, y = fitted(mP5), col = ConifWet)) +
    geom_point() +
    geom_smooth() +
    theme_light()
```

Final battle of Poisson models: 

```{r}
MuMIn::model.sel(mP0, mP1, mP12, mP13, mP14, mP2, mP3, mP4, mP5)
```

Of course, the most complex model wins but the Chi-square test is still 
significant (indicating lack of fit). Let's try different error distribution.

## Different error distributions

We will use the 2-variable model with interaction:

```{r}
mP <- glm(Count ~ Decid * ConifWet, data = x, family = poisson)
summary(mP)
```

Let us try the Negative Binomial distribution first. This distribution is 
related to Binomial experiments (number of trials required to get a fixed 
number of successes given a binomial probability). It can also be derived as a 
mixture of Poisson and Gamma distributions (see [Wikipedia](https://en.wikipedia.org/wiki/Negative_binomial_distribution#Gamma%E2%80%93Poisson_mixture)), which is a kind 
of hierarchical model. In this case, the Gamma distribution acts as an i.i.d. 
random effect for the intercept:
$Y_i\sim Poisson(\lambda_i)$, $\lambda_i  \sim Gamma(e^{\beta_0+\beta_1 x_{1i}}, \gamma)$, 
where $\gamma$ is the Gamma variance.

The Negative Binomial variance (using the parametrization common in R functions) 
is a function of the mean and the scale: $V(\mu) = \mu + \mu^2/\theta$.

```{r}
mNB <- MASS::glm.nb(Count ~ Decid * ConifWet, data = x)
summary(mNB)
```

Next, we look at zero-inflated models. In this case, the mixture distribution 
is a Bernoulli distribution and a count distribution (Poisson or Negative Binomial, 
for example). The 0's can come from both the zero and the count distributions, 
whereas the >0 values can only come from the count distribution: 
$A_i \sim Bernoulli(\varphi)$, $Y_i \sim Poisson(A_i \lambda_i)$.

The zero part of the zero-inflated models are often parametrized as probability 
of zero ($1-\varphi$), as in the `pscl::zeroinfl` function:

```{r}
## Zero-inflated Poisson
mZIP <- pscl::zeroinfl(Count ~ Decid * ConifWet | 1, x, dist = "poisson")
summary(mZIP)

## Zero-inflated Negative Binomial
mZINB <- pscl::zeroinfl(Count ~ Decid * ConifWet | 1, x, dist = "negbin")
summary(mZINB)
```

```{r}
## Hurdle Poisson
mHP <- pscl::hurdle(Count ~ Decid * ConifWet | 1, x, dist = "poisson")
summary(mHP)

## Hurdle Negative Binomial
mHNB <- pscl::hurdle(Count ~ Decid * ConifWet | 1, x, dist = "negbin")
summary(mHNB)
```

_Exercise_: compare ZI and Hurdle coefficients. Anything surprising? What do the coefficients for the zero part of the model mean?

Now we compare the four different parametric models:

```{r}
MuMIn::model.sel(mP, mNB, mZIP, mZINB, mHP, mHNB)
```

Our best model is the ZINB. The probability of observing a zero as part of the 
zero distribution is back transformed from the zero coefficient using the inverse logit function:

```{r}
unname(plogis(coef(mZIP, "zero"))) # P of 0 (not 1!)
```

## Mixed models

It is also common practice to consider generalized linear mixed models (GLMMs) 
for count data. These mixed models are usually considered as Poisson-Lognormal 
mixtures. The simplest, so called i.i.d., case is similar to the Negative Binomial, 
but instead of Gamma, we have Lognormal distribution: 
$Y_i\sim Poisson(\lambda_i)$, 
$log(\lambda_i) = \beta_0+\beta_1 x_{1i}+\epsilon_i$, $\epsilon_i \sim Normal(0, \sigma^2)$, 
where $\sigma^2$ is the Lognormal variance on the log scale.

We can use the `lme4::glmer` function: use `SiteID` as random effect (we have exactly $n$ random effects).

```{r}
mPLN1 <- lme4::glmer(Count ~ Decid * ConifWet + (1 | SiteID),
    data = data.frame(SiteID = rownames(x), x), family = poisson
)
summary(mPLN1)
```

Note: the number of unknowns we have to somehow estimate is now more than the 
number of observations we have. How is that possible?

Alternatively, we can use `SurveyArea` as a grouping variable. We have now 
$m < n$ random effects, and survey areas can be seen as larger landscapes 
within which the sites are clustered: $Y_ij\sim Poisson(\lambda_ij)$, 
$log(\lambda_ij) = \beta_0+\beta_1 x_{1ij}+\epsilon_i$, $\epsilon_i \sim Normal(0, \sigma^2)$. 
The index $i$ ($i=1,...,m$) defines the cluster (survey area), the 
$j$ ($j=1,...,n_i$) defines the sites within survey area $i$ ($n = \sum_{i=1}^m n_i$).

```{r}
mPLN2 <- lme4::glmer(Count ~ Decid * ConifWet + (1 | SurveyArea),
    data = data.frame(SurveyArea = detect::josm$surveys$SurveyArea, x),
    family = poisson
)
summary(mPLN2)
```

In the battle of distributions (keeping the linear predictor part the same) 
the clustered GLMM was best supported:

```{r}
tmp <- AIC(mP, mNB, mZIP, mZINB, mHP, mHNB, mPLN1, mPLN2)
tmp$delta_AIC <- tmp$AIC - min(tmp$AIC)
tmp[order(tmp$AIC), ]
```

# Summary

- Quantification of uncertainty is the hallmark of statistical inference.
- Prediction uncertainty has two major components: Estimation
uncertainty and Intrinsic variability.
- Estimation uncertainty can be reduced by increasing the sample size
and good design.
- Intrinsic variability is usually the substantial part of the prediction
uncertainty.
- Intrinsic variability can be reduced by using auxiliary information
through regression modelling.
- One should use appropriate statistical model to model intrinsic
variability.
