---
title: "Understanding observation error"
subtitle: "Point count data analysis workshop 2025"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
# options(width = 53, scipen = 999)
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
})
set.seed(1234)
```

# Observation error

How good is our measurement?

- Accuracy: how close is the measurement to the true value?
- Precision: how close are repeated measurements to each other?

When we talk about error in variables, we usually refer to accuracy. 
Most often we assume that the error is so small that we can ignore it,
or that the error is random and it'll just contribute to the overall variability.
We can think of many sources of error in our data, e.g. due to spatial positioning,
or variables being derived from satellite imagery (older vintage, cloud cover, etc.).

When it comes to the response variable, we often have to deal with observation error. 
This is the case when we are trying to measure something that is not directly observable, 
e.g. occupancy status or abundance of a species. We can only get an estimate of the 
true value, and this estimate is subject to observation error. 

In the simplest case, say, we are interested in whether a site is occupied by a species or not. 
We can only observe the species if it is present (true positive), 
and we might fail to detect it even if it is present (false negative). 
The other case is when the species is absent, but the observer mistakenly thinks it is present (false positive),
e.g. due to misidentification. In most situations, we assume that there are no false positives, and we only deal with false negatives.

- If the true occupancy status is 0 (absent), we will always observe 0
- If the true occupancy status is 1 (present), we will observe 1 or 0, this is the detection error we need to account for in our models

# Introduction to simulations

The goal is to implement

- a data generating mechanism
- using random numbers

Here is the most basic simulation: the coin flip.

We assume that the coin is fair, therefore:

- the probability of getting head ($y=1$) is $p=0.5$
- the probability of getting tail ($y=0$) is $1-p=0.5$

Here is code for setting the probability value $p$
and getting the outcome $y$ using Uniform random numbers:

1. We generate a random number (`u`) between 0 and 1
2. The outcome is 1 if the number is $<p$
3. The outcome is 0 if the random number of $\ge p$

```{r}
p <- 0.5

u <- runif(n = 1, min = 0, max = 1)
u

y <- ifelse(u < p, 1, 0)
y
```

::: {.callout-tip}
## Reproducibility with random numbers

Every time you run the code above, you'll get a different value for `u`
and possibly a different outcome. To make this reproducible,
we can set the random seed with e.g. `set.seed(123)`.
:::

The `runif()` function takes 3 arguments:

- `n` is the number of trials, i.e. number of coin flips
- `min` and `max` define the range, here we used the unit range $0-1$

Here is the histogram of a $Uniform(0, 1)$ random variable:

```{r}
u <- runif(n = 100, min = 0, max = 1)
summary(u)
hist(u)
```

We use the random numbers to repeat the coin flip experiment 100 times
and tabulate the results:

```{r}
ifelse(u < p, 1, 0) |>
    table(y = _) |>
    as.data.frame() |>
    dplyr::mutate(Prop = Freq / sum(Freq))
```

We will see uses of the Uniform distribution later.

Another way to run the coin flip experiment in R is to use the Bernoulli
distribution.

```{r}
y <- rbinom(n = 100, size = 1, prob = p)
summary(y)
table(y = y) |>
    as.data.frame() |>
    dplyr::mutate(Prop = Freq / sum(Freq))
```

The `rbinom()` function takes 3 arguments:

- `n` is the number of observations
- `size` is the number of trials
- `prob` is the probability of the outcome to be 1

The `rbinom` name refers to the Binomial distribution, which is multiple
independent Bernoulli trials. This is what the `size` argument refers to,
i.e. `size = 1` means Bernoulli.
