---
title: "Fitting occupancy models"
subtitle: "Analysing Ecological Data with Detection Error"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
# options(width = 53, scipen = 999)
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(unmarked)
    library(detect)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
})
set.seed(1234)
```

# Simulating occupancy data

Let's see how we can use the Bernoulli distribution and the `rbinom`
function to simulate occupancy data.

Let us sample $n=50$ locations and observe the occupancy status of
Ovenbirds at each of the locations:

- The true occupancy status at the $i$th location is $W_{i}$.
- The probability of occupancy is $P(W_{i}=1)=\phi$.

The true occupancy probability $\phi$ is denoted with `phi.true`:

```{r}
n <- 50
phi.true <- 0.4

W <- rbinom(n = n, size = 1, prob = phi.true)

table(W = W)
```


There are many reasons why we might not observe the true occupancy status $W$.
When the data that we collect ($Y$) is not the true occupancy status $W$, 
we say that there is detection error. It can manifest the following way:

We can capture detection error also as a probability $p$ which
is the probability of detecting the species when present ($W_i=1$).

- If the species is absent ($W_i=0$), we will always observe 0, thus $Y_i=0$: $P(Y_i=0|W_{i}=0)=1$.
- If $W_i=1$, we might observe 0 (missed all the birds) or 1:
  - species detected when present, $P(Y_i=1|W_{i}=1)=p$
  - species not detected when present, $P(Y_i=0|W_{i}=1)=1-p$

Let's simulate what we observe, the `Y` vector.
We use the `rbinom()` function with `n = n` as before,
but the size argument is not a fixed 1, but the true status `W`.
When `W` is 0, the function will always return 0,
when `W` is 1, the function will reyrun 1 or 0 according the the
detection probability $p$, denoted as `p.true`:

```{r}
p.true <- 0.6
Y <- rbinom(n = n, size = W, prob = p.true)

table(W = W, Y = Y)
```

The cross-tabulation show how many times we observed 1's vs 0's when the
true status was 1.

Using the observations as if those would represent the true status
is called the _naive_ approach. The naive approach assume that there is
no observation error:

```{r}
mean(Y)
```

But we see that the mean of `Y` is quite far from `phi.true`.

Fitting a logistic regression (aka Binomial GLM) to the data yields the
same result:

```{r}
m1 <- glm(Y ~ 1, family = binomial)
coef(m1)
plogis(coef(m1))
```

The coefficient for the intercept is the maximum likelihood estimate on the
logit scale. To transform that to the $0-1$ probability scale, we can use
the `plogis()` function that implements the inverse logistic transformation.

The unobserved `W` variable is often called a _latent_ variable.
It is latent in the sense that it cannot be directly observed.
We need to find ways to:

- use survey design to minimize the detection error, or
- use statistical models to correct for detection error, or
- the combination of the 2.

# Simulating multiple visits

A feature of multiple-visit methods is that we visit the same site $T$ times.
Each visit ($t=1,\ldots,T$) gives us a different observation, $Y_{it}$.

Key assumptions are the following:

- the visits are independent of each other
- the true status $W$ stays the same over the visits ($W_i=W_{it}$)

We can generalize the simulation code to any number of `T` visits as:

```{r}
T <- 5
Y <- matrix(NA, n, T)
for (t in 1:T) {
    Y[, t] <- rbinom(n = n, size = W, prob = p.true)
}
```

A more concise way of writing the above code without a loop is:

```{r}
#| eval: false
Y <- replicate(T, rbinom(n = n, size = W, prob = p.true))
```

If we inspect any row (site) from the `Y` matrix, we see a series of 1's and 0's
for sites where the species is present. This is called the _detection history_.

```{r}
data.frame(
    W = head(W[W > 0]),
    Visit = head(Y[W > 0, ])
)
```

The detection history is all 0's for sites where the species is absent:

```{r}
data.frame(
    W = head(W[W == 0]),
    Visit = head(Y[W == 0, ])
)
```

We can compare the maximum of the observed values over the visits at each site:

```{r}
Y_max <- apply(Y, 1, max)
table(W = W, Y_max = Y_max)
```

Fit the logistic regression model to `Y_max`:

```{r}
m2 <- glm(Y_max ~ 1, family = binomial)
coef(m2)
plogis(coef(m2))
```

::: {.callout-tip}
## What happens when we increase the number of visits?

Change the value of `T` and compare the `Y_max` to `W`. 
What happens to our _naive_ estimator of using `Y_max`?
:::

The maximum over a large number of visits will approach the latent
variable `W`:

```{r}
n2 <- 500
W2 <- rbinom(n = n2, size = 1, prob = phi.true)
Tvals <- 1:10

data.frame(
    T = Tvals,
    Y_max = sapply(Tvals, \(T) {
        mean(apply(
            replicate(T, rbinom(n = n2, size = W, prob = p.true)),
            1, max
        ))
    })
) |> ggplot(aes(x = T, y = Y_max)) +
    geom_hline(yintercept = phi.true, lty = 2, col = "red") +
    geom_hline(yintercept = mean(W), lty = 2, col = "blue") +
    geom_line() +
    ylim(0, 1) +
    scale_x_continuous(breaks = Tvals) +
    theme_light()
```

# Fitting occupancy models

The unmarked package implements the multiple-visit occupancy model.

::: {.callout-caution collapse="true"}
## References

MacKenzie et al., 2002.
Estimating site occupancy rates when detection probabilities are less than one.
Ecology, 83:2248--2255.
[Fulltext](https://www.sfu.ca/~lmgonigl/materials-qm/papers/mackenzie-2002-2248.pdf),
DOI 10.1890/0012-9658(2002)083[2248:ESORWD]2.0.CO;2
:::

- Organize the data and an unmarked occupancy data frame
- Fit the model with the `occu`
- the `~1 ~1` formula says that we do not have covariates, just the intercepts

```{r}
umf <- unmarked::unmarkedFrameOccu(y = Y)
summary(umf)

m3 <- unmarked::occu(~1 ~ 1, umf)
m3
```

We use the `plogis()` function again to transform the estimates to the
probability scale and compare with our `phi.true` and `p.true` values:

```{r}
plogis(coef(m3, type = "det"))
plogis(coef(m3, type = "state"))
```

## Maximum likelihood estimator

The `glm()` and `occu()` functions use maximum likelihood to find the
parameter estimates for a given data set.
In other words, we coefficients (maximum likelihood estimate, or MLE)
maximize the likelihood function for the data set in question.
The likelihood function can be relatively simple and easy to calculate,
or it can be computationally challenging to compute (e.g. for hierarchical
or mixed models).

The likelihood function $L$ for the simple occupancy model with 
parameters $p$ and $\phi$ for multiple visits data can be written as:

$$L(p, \phi; y_{1,1}, \ldots, y_{n,T})  = \prod_{i=1}^{n} \left[ \phi \left( \binom{T}{y_{i \cdot}} p^{y_{i \cdot}} (1 - p)^{T - y_{i \cdot}} \right) + (1 - \phi) I(y_{i \cdot} = 0)\right]$$

where $y_{i \cdot} = \sum^{t=1}_{T} y_{i,t}$ and 
$I( y_{i \cdot} = 0 )$ is an indicator function that is equal to 1 if $y_{i \cdot} = 0$.

Here is the R code to calculate the log likelihood for the occupancy model:

```{r}
L_fun_occu <- function(Y, p, phi) {
    ydot <- rowSums(Y)
    T <- ncol(Y)
    L <- prod(
        phi *
            (choose(T, ydot) * p^ydot * (1 - p)^(T - ydot)) +
            (1 - phi) * (ydot == 0)
    )
    L
}
```

Next, we evaluate the likelihood function at different values of `p` and `phi`
while keeping the data `Y` constant. We set up the grid for this using
`expand.grid()`:

```{r}
g <- 100
grid <- expand.grid(
    p = seq(0, 1, length.out = g),
    phi = seq(0, 1, length.out = g),
    L = NA
)

for (i in 1:nrow(grid)) {
    grid$L[i] <- L_fun_occu(
        Y = Y,
        p = grid$p[i],
        phi = grid$phi[i]
    )
}
```

When we plot the likelihood surface, we see the maximum, the lines indicate
the true probability values:

```{r}
image(
    list(
        x = unique(grid$p),
        y = unique(grid$phi),
        z = matrix(grid$L, g, g)
    ),
    xlab = "p",
    ylab = expression(varphi)
)
abline(h = phi.true, v = p.true, col = 1, lwd = 1)

grid |> ggplot(aes(x = p, y = phi, z = L)) +
    geom_contour_filled(show.legend = FALSE) +
    geom_hline(yintercept = phi.true) +
    geom_vline(xintercept = p.true) +
    xlab("p") +
    ylab(expression(phi)) +
    theme_light()

grid[which.max(grid$L), ]
```

If the rgl package is installed, we can view the surface in 3D:

```{r}
if (interactive()) {
    library(rgl)

    L_mat <- matrix(grid$L, g, g)
    dcpal_grbu <- colorRampPalette(c("#18bc9c", "#3498db"))
    Col <- rev(dcpal_grbu(12))[cut(L_mat, breaks = 12)]

    open3d()
    persp3d(
        x = unique(grid$p),
        y = unique(grid$phi),
        z = L_mat / max(L_mat),
        col = Col,
        theta = 50, phi = 25, expand = 0.75, ticktype = "detailed",
        xlab = "p", ylab = expression(phi), zlab = "L"
    )
    quads3d(
        x = rep(p.true, 4),
        y = c(0, 0, 1, 1),
        z = c(0, 1, 1, 0),
        alpha = 0.5, col = 2
    )
    quads3d(
        x = c(0, 0, 1, 1),
        y = rep(phi.true, 4),
        z = c(0, 1, 1, 0),
        alpha = 0.5, col = 4
    )
}
```

::: {.callout-note}
We will circle back to these plots later, feel free to explore it
with different settings of $n$, $T$, $p$, and $\phi$.
:::

# Field data analysis

We will use the ABMI data, which has 3 equal (3.3 min) time intervals for each point count visit.
We use the point level data, but only for the center point, so we can keep 
sample size small and run the code quickly. When analyzing such data sets
for real, we would use all the data and not subset it like this.

```{r}
if (!interactive()) {
    d <- readRDS("../data/ab_birds_site_year_pt.rds")
} else {
    d <- readRDS("data/ab_birds_site_year_pt.rds")
}

spp <- "yellowrumped_warbler"

x <- data.frame(
    y1 = ifelse(d$visits$visit1[, spp] > 0, 1, 0),
    y2 = ifelse(d$visits$visit2[, spp] > 0, 1, 0),
    y3 = ifelse(d$visits$visit3[, spp] > 0, 1, 0),
    n1 = d$visits$visit1[, spp],
    n2 = d$visits$visit2[, spp],
    n3 = d$visits$visit3[, spp],
    d$vars
)
x$yday <- as.POSIXlt(x$fielddate)$yday / 365
x$hr <- x$fieldtime / 24
x$y_max <- pmax(x$y1, x$y2, x$y3)
x$n_max <- pmax(x$n1, x$n2, x$n3)
x <- x |>
    filter(
        hr > 0.1,
        hr < 0.6,
        yday > 0.37,
        yday < 0.5
    )
x0 <- x[x$point == 1, ]
```

Naive occupancy estimates for the 3 visits:

```{r}
m_v1 <- glm(y1 ~ 1, data = x0, family = binomial)
coef(m_v1)
plogis(coef(m_v1))
mean(x0$y1)

m_v2 <- glm(y2 ~ 1, data = x0, family = binomial)
coef(m_v2)
plogis(coef(m_v2))
mean(x0$y2)

m_v3 <- glm(y3 ~ 1, data = x0, family = binomial)
coef(m_v3)
plogis(coef(m_v3))
mean(x0$y3)
```

Given that the 3 subsequent 'visits' are actually time intervals within the same point count,
we can safely assume that the true occupancy status does not change between the visits
(closed population assumption). Therefore, we can use the `y_max` variable to fit a naive occupancy model
to approximate the true occupancy status:

```{r}
m_max <- glm(y_max ~ 1, data = x0, family = binomial)
coef(m_max)
plogis(coef(m_max))
mean(x0$y_max)
```

_Question_: What do you think about the closed population assumption?
What if visits are days, weeks, months, or years apart?

Let's fit the occupancy model to the data using all 3 visits:

```{r}
umf <- unmarked::unmarkedFrameOccu(y = x0[, c("y1", "y2", "y3")])
summary(umf)

m_mv <- unmarked::occu(~1 ~ 1, umf)
m_mv

plogis(coef(m_mv, type = "det"))
plogis(coef(m_mv, type = "state"))
```

Use covariates: for status and detection

```{r}
umf2 <- unmarked::unmarkedFrameOccu(
    y = x0[, c("y1", "y2", "y3")],
    siteCovs = x0[, c("pconif", "thf")],
    obsCovs = list(
        hr = x0[, c("hr", "hr", "hr")],
        yday = x0[, c("yday", "yday", "yday")]
    )
)
summary(umf2)

m_mv2 <- unmarked::occu(
    ~ hr + I(hr^2) + yday ~ pconif + I(pconif^2) + thf,
    umf2
)
m_mv2

m_mv@AIC
m_mv2@AIC
```

Fitted values reflect expected values of the marginal distribution,
i.e. combining the state and detection processes:

```{r}
summary(fitted(m_mv2))

colMeans(umf2@y)
```

We can calculate the expected occupancy probability for each site by using the coefficients of the model matrix:

```{r}
f <- m_mv2@formula[-2]
f
X <- model.matrix(f, x0)
head(X)

coef(m_mv2, type = "state")
X[1, ] * coef(m_mv2, type = "state")
sum(X[1, ] * coef(m_mv2, type = "state"))
head(X %*% coef(m_mv2, type = "state"), 1)
```

```{r}
fit_mv2 <- plogis(X %*% coef(m_mv2, type = "state"))
summary(fit_mv2)
hist(fit_mv2)
```

Visualize the fitted occupancy probabilities with respect to the covariates:

```{r}
data.frame(x0, p_occ = fit_mv2) |>
    ggplot(aes(x = pconif, y = p_occ, col = thf)) +
    geom_point() +
    ylim(0, 1)

data.frame(x0, p_occ = fit_mv2) |>
    ggplot(aes(x = as.factor(y_max), y = p_occ)) +
    geom_boxplot() +
    ylim(0, 1)

m_max2 <- glm(y_max ~ pconif + I(pconif^2) + thf, data = x0, family = binomial)

data.frame(p_occ = fit_mv2, p_max = fitted(m_max2)) |>
    ggplot(aes(x = p_max, y = p_occ)) +
    geom_point() +
    geom_abline(lty = 2) +
    geom_smooth(method = "lm", se = FALSE)

mean(fit_mv2 / fitted(m_max2))
```

Predict detection probability given detection covariates:

```{r}
Z <- model.matrix(m_mv2@formula[-3], x0)
head(Z)
det_mv2 <- plogis(Z %*% coef(m_mv2, type = "det"))
summary(det_mv2)
hist(det_mv2)
```

```{r}
data.frame(x0, p_det = det_mv2) |>
    ggplot(aes(x = yday, y = p_det, col = hr)) +
    geom_point() +
    ylim(0, 1)
```

_Question_: How would we assess model fit? Can we use AUC/ROC?

What do we get when only using 2 visits?

```{r}
umf_v2 <- unmarked::unmarkedFrameOccu(y = x0[, c("y1", "y2")])
m_mv_v2 <- unmarked::occu(~1 ~ 1, umf_v2)
m_mv_v2
plogis(coef(m_mv_v2, type = "det"))
plogis(coef(m_mv_v2, type = "state"))
```

How about only using 1 visit?

```{r}
umf_v1 <- unmarked::unmarkedFrameOccu(y = x0[, c("y1"), drop = FALSE])
m_mv_v1 <- unmarked::occu(~1 ~ 1, umf_v1)
m_mv_v1
plogis(coef(m_mv_v1, type = "det"))
plogis(coef(m_mv_v1, type = "state"))
```

Something is wrong here, right?
Let put this aside for now, we will come back to this later.
