---
title: "Fitting N-mixture models"
subtitle: "Point count data analysis workshop 2025"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
# options(width = 53, scipen = 999)
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(unmarked)
    library(detect)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
})
set.seed(1234)
```

# Simulating count data

Simulating counts is similar to occupancy. We need a count distribution.
The most basic count distribution is Poisson which has 1 parameter,
$\lambda$, which is the mean. The variance also happens to equal the mean.
Use the `rpois()` function in R to generate random numbers from the
poisson distribution. Here, $N_i$ ($i=1,\ldots,n$) is the abundance 
(number of individuals) at the $i$th location.

```{r}
n <- 50
n2 <- 500
p.true <- 0.6

lambda.true <- 4.2

N <- rpois(n = n, lambda = lambda.true)

table(N)
summary(N)

table(N) |> plot(ylab = "Frequency")
```

# Observation error for counts

Here is how we introduce observation error to count data:

- select a location,
- take the count $N_i$,
- for each individual ($1, 2, \ldots, N_i$), use the Bernoulli distribution with probability $p$
  to determine if the individual was detected (1) or not (0)

The number of individuals detected ($Y_i$) is less than or equal to $N_i$:
$Y_i \le N_i$. For example, if $N_i = 4$, $Y_i$ can be 0, 1, 2, 3, or 4.

To express things mon concisely, we can also use a Binomial distribution
with $N_i$ as the number of trials and probability $p$:

```{r}
Y <- rbinom(n = n, size = N, prob = p.true)

print(table(N = N, Y = Y), zero.print = ".")
```

# Count data with multiple visits

When visiting each site $T$ times, the observed data will be organized in a
$n$ by $T$ matrix as how we saw it for occupancy:

```{r}
Y <- matrix(NA, n, T)
for (t in 1:T) {
    Y[, t] <- rbinom(n = n, size = N, prob = p.true)
}

# alternatively
# Y <- replicate(T, rbinom(n = n, size = N, prob = p.true))
```

Multiple-visit count model have similar assumptions as occupancy models,
most importantly that $N_{it}=N_i$. This condition is also referred to as
the closed population assumption, i.e. there is no immigration,
emigration, birth, or death between visits.

```{r}
data.frame(N = N, Visit = Y) |> head()
```

As we saw before, we can use the `Y_max` to fit the naive count model,
but instead of the `plogis()` function, we use `log()` as the inverse of the
logarithmic link used for Poisson GLM:

```{r}
Y_max <- apply(Y, 1, max)
m4 <- glm(Y_max ~ 1, family = poisson)
coef(m4)
exp(coef(m4))
```

With enough visits, tha `Y_max` will approach the  latent variable `N`:

```{r}
N2 <- rpois(n = n2, lambda = lambda.true)
Tvals <- c(1, 2, 4, 8, 10, 15, 20, 30, 40, 50)

data.frame(
    T = Tvals,
    Y_max = sapply(Tvals, \(T) {
        mean(apply(
            replicate(T, rbinom(n = n, size = N, prob = p.true)),
            1, max
        ))
    })
) |> ggplot(aes(x = T, y = Y_max)) +
    geom_hline(yintercept = lambda.true, lty = 2, col = 2) +
    geom_hline(yintercept = mean(N), lty = 2, col = 4) +
    geom_line() +
    ylim(0, NA) +
    scale_x_continuous(breaks = Tvals) +
    theme_light()
```


# Fitting N-mixture models

Fit the multiple-visit count model, the so called N-mixture model,
using the `pcount()` function of the unmarked R package.

The `unmarked::unmarkedFramePCount()` function is used to organize
the count data:

```{r}
umf <- unmarked::unmarkedFramePCount(y = Y)
summary(umf)

m5 <- unmarked::pcount(~1 ~ 1, umf, K = 50)
m5

plogis(coef(m5, type = "det"))
exp(coef(m5, type = "state"))
```

::: {.callout-caution collapse="true"}
## References

Royle, 2004.
N-mixture models for estimating population size from spatially replicated counts.
Biometrics, 60:108--115.
[Fulltext](https://ecology.ghislainv.fr/publications/biblio/Royle2004-Biometrics.pdf),
DOI 10.1111/j.0006-341X.2004.00142.x
:::

The `K` argument of the `pcount()` function is the upper index of integration 
for N-mixture. This will make more sense once we take a look at the
likelihood function in the next section.

## N-mixture likelihood

The likelihood function for the N-mixture model can be written as:

$$
L(p, \lambda; y_{1,1}, \ldots, y_{n,T}) = 
\prod^{n}_{i=1}
\sum^{K}_{N_{i}=Y_{max,i}}
\prod^{T}_{t=1}
e^{-\lambda_{i}}\frac{\lambda_{i}^{N_{i}}}{N_{i}!}\left(\begin{array}{c}
N_{i}\\
Y_{it}
\end{array}\right)
p^{Y_{it}} (1-p)^{N_{i}-Y_{it}}
$$

This distribution is a mixture of Poisson and Binomial distributions.

The $K$ value strictly speaking should be $\infty$, but the estimates
won't change much as long as $K$ is large enough relative to the maximum of
$N_i$. Of course for real data we do not know what the values of $N_i$ are.
The numerical integration (summation) for site $i$ goes
from $Y_{max,i}$ (`Y_max`) to $K$, because we know that $Y_{it} \le N_i$.

We can write this as a log likelihood function:

```{r}
logL_fun_pcount_R <- function(Y, p, lambda, n, T, Y_max, K) {
    L <- rep(NA, n)
    for (i in 1:n) {
        S <- 0
        for (Nit in Y_max[i]:K) {
            v <- 0
            for (j in 1:T) {
                v <- v + dbinom(Y[i, j], Nit, p, log = TRUE) +
                    dpois(Nit, lambda, log = TRUE)
            }
            S <- S + exp(v)
        }
        L[i] <- S
    }
    sum(log(L))
}

# using C code from unmarked - much faster
logL_fun_pcount_C <- function(Y, p, lambda, n, T, Y_max, K = 50) {
    nll <- unmarked:::nll_pcount(
        beta = c(log(lambda), qlogis(p)),
        n_param = c(1, 1, 0),
        y = Y,
        X = matrix(1, n, 1),
        V = matrix(1, n * T, 1),
        X_offset = rep(0, n),
        V_offset = matrix(1, n * T, 1),
        K = K,
        Kmin = Y_max,
        mixture = 1,
        threads = 1
    )
    -nll
}
```

```{r}
g <- 50
grid <- expand.grid(
    p = seq(0, 1, length.out = g),
    lambda = seq(0, lambda.true * 2, length.out = g),
    logL = NA
)

logL_fun_pcount <- logL_fun_pcount_C
for (i in 1:nrow(grid)) {
    grid$logL[i] <- logL_fun_pcount(
        Y = Y,
        p = grid$p[i],
        lambda = grid$lambda[i],
        n = n, T = T, Y_max = Y_max, K = 50
    )
}
grid$logL[is.infinite(grid$logL)] <- NA
```

The images showing the likelihood surface:

```{r}
image(
    list(
        x = unique(grid$p),
        y = unique(grid$lambda),
        z = matrix(exp(grid$logL), sqrt(nrow(grid)))
    ),
    xlab = "p",
    ylab = expression(lambda)
)
abline(h = lambda.true, v = p.true, col = 1, lwd = 1)

grid |> ggplot(aes(x = p, y = lambda, z = exp(logL))) +
    geom_contour_filled(show.legend = FALSE) +
    geom_hline(yintercept = lambda.true) +
    geom_vline(xintercept = p.true) +
    xlab("p") +
    ylab(expression(lambda)) +
    theme_light()
grid[which.max(grid$logL), ]
```

::: {.callout-note}
We will circle back to these plots later, feel free to explore it
with different settings of $n$, $T$, $p$, and $\lambda$.
:::

# Field data analysis

We will use the ABMI data, which has 3 equal (3.3 min) time intervals for each point count visit.
We use the point level data, but only for the center point, so we can keep 
sample size small and run the code quickly. When analyzing such data sets
for real, we would use all the data and not subset it like this.

```{r}
if (!interactive()) {
    d <- readRDS("../data/ab_birds_site_year_pt.rds")
} else {
    d <- readRDS("data/ab_birds_site_year_pt.rds")
}

spp <- "yellowrumped_warbler"

x <- data.frame(
    y1 = ifelse(d$visits$visit1[, spp] > 0, 1, 0),
    y2 = ifelse(d$visits$visit2[, spp] > 0, 1, 0),
    y3 = ifelse(d$visits$visit3[, spp] > 0, 1, 0),
    n1 = d$visits$visit1[, spp],
    n2 = d$visits$visit2[, spp],
    n3 = d$visits$visit3[, spp],
    d$vars
)
x$yday <- as.POSIXlt(x$fielddate)$yday / 365
x$hr <- x$fieldtime / 24
x$y_max <- pmax(x$y1, x$y2, x$y3)
x$n_max <- pmax(x$n1, x$n2, x$n3)
x <- x |>
    filter(
        hr > 0.1,
        hr < 0.6,
        yday > 0.37,
        yday < 0.5
    )
x0 <- x[x$point == 1, ]
```

Naive abundance estimates for the 3 visits:

```{r}
m_v1 <- glm(n1 ~ 1, data = x0, family = poisson)
coef(m_v1)
exp(coef(m_v1))
mean(x0$n1)

m_v2 <- glm(n2 ~ 1, data = x0, family = poisson)
coef(m_v2)
exp(coef(m_v2))
mean(x0$n2)

m_v3 <- glm(n3 ~ 1, data = x0, family = poisson)
coef(m_v3)
exp(coef(m_v3))
mean(x0$n3)
```

Use the maximum count to fit the naive model:

```{r}
m_max <- glm(n_max ~ 1, data = x0, family = poisson)
coef(m_max)
exp(coef(m_max))
mean(x0$n_max)
```

N-mixture model using the `pcount()` function:

```{r}
# here N-mix version using pcount
umf <- unmarked::unmarkedFramePCount(y = x0[, c("n1", "n2", "n3")])
summary(umf)

m_mv <- unmarked::pcount(~1 ~ 1, umf, K = 100)
m_mv

plogis(coef(m_mv, type = "det"))
exp(coef(m_mv, type = "state"))
```

Use covariates to model abundance and detection:

```{r}
umf2 <- unmarked::unmarkedFramePCount(
    y = x0[, c("n1", "n2", "n3")],
    siteCovs = x0[, c("pconif", "thf")],
    obsCovs = list(
        hr = x0[, c("hr", "hr", "hr")],
        yday = x0[, c("yday", "yday", "yday")]
    )
)
summary(umf2)

m_mv2 <- unmarked::pcount(
    ~ hr + I(hr^2) + yday ~ pconif + I(pconif^2) + thf,
    umf2,
    K = 100
)
m_mv2

m_mv@AIC
m_mv2@AIC
```

```{r}
summary(fitted(m_mv2))
colMeans(umf2@y)

X <- model.matrix(m_mv2@formula[-2], x0)
fit_mv2 <- exp(X %*% coef(m_mv2, type = "state"))
hist(fit_mv2)
summary(fit_mv2)

data.frame(x0, lam = fit_mv2) |>
    ggplot(aes(x = pconif, y = lam, col = thf)) +
    geom_point() +
    ylim(0, NA)

data.frame(x0, lam = fit_mv2) |>
    ggplot(aes(x = as.factor(n_max), y = lam)) +
    geom_boxplot() +
    ylim(0, NA)

m_max2 <- glm(n_max ~ pconif + I(pconif^2) + thf, data = x0, family = poisson)

data.frame(lam = fit_mv2, n_max = fitted(m_max2)) |>
    ggplot(aes(x = n_max, y = lam)) +
    geom_point() +
    geom_abline(lty = 2) +
    geom_smooth(method = "lm", se = FALSE)
mean(fit_mv2 / fitted(m_max2))

Z <- model.matrix(m_mv2@formula[-3], x0)
head(Z)
det_mv2 <- plogis(Z %*% coef(m_mv2, type = "det"))
summary(det_mv2)
hist(det_mv2)

data.frame(x0, p_det = det_mv2) |>
    ggplot(aes(x = yday, y = p_det, col = hr)) +
    geom_point() +
    ylim(0, 1)
```

What do we get when only using 2 visits?

```{r}
umf_v2 <- unmarked::unmarkedFramePCount(y = x0[, c("n1", "n2")])
m_mv_v2 <- unmarked::pcount(~1 ~ 1, umf_v2, K = 100)
m_mv_v2
plogis(coef(m_mv_v2, type = "det"))
exp(coef(m_mv_v2, type = "state"))
```

How about only using 1 visit?

```{r}
umf_v1 <- unmarked::unmarkedFramePCount(y = x0[, c("n1"), drop = FALSE])
m_mv_v1 <- unmarked::pcount(~1 ~ 1, umf_v1, K = 100)
m_mv_v1
plogis(coef(m_mv_v1, type = "det"))
exp(coef(m_mv_v1, type = "state"))
```

Single visit estimates for the constant abundance model are not reliable.

There are tools out there (e.g. data cloning) that you can used to see
if the model you are fitting is identifiable or not.

_Exercise_: Try `"P"`, `"NB"`, `"ZIP"` mixtures for `pcount()`.
