---
title: "Fitting N-mixture models"
subtitle: "Analysing Ecological Data with Detection Error"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
# options(width = 53, scipen = 999)
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(unmarked)
    library(detect)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
})
set.seed(1234)
```

Following simulation code from our MEE paper:
<https://github.com/psolymos/detect/blob/master/extras/revisitingSV/simuls6.R>

```{r}
n <- 200 # number of sites
x1 <- runif(n, 0, 1) # random predictors
x2 <- rnorm(n, 0, 1)
x3 <- runif(n, 0, 1)
x4 <- rnorm(n, 0, 1)
x <- data.frame(x1 = x1, x2 = x2, x3 = x3, x4 = x4)

X <- model.matrix(~ x1 + x2, x)
Z <- model.matrix(~ x3 + x4, x)

K <- 100
T <- 3

## true coefficients
beta <- c(2, -0.8, 0.6)
theta <- c(-1, 2, -0.5)

omega <- 0.999 # apparent survival probability in [0,1]

p <- plogis(drop(Z %*% theta))
lambda <- exp(drop(X %*% beta))
```

Dail & Madsen's `pcountOpen()` implements the following population dynamics models:

- "constant" indicates that there is no relationship between omega and gamma. 
- "autoreg" is an auto-regressive model in which recruitment is modeled as `gamma*N[i,t-1]`. 
- "notrend" model gamma as `lambda*(1-omega)` such that there is no temporal trend. 
- "trend" is a model for exponential growth, `N[i,t] = N[i,t-1]*gamma`, where gamma in this case is finite rate of increase (normally referred to as lambda). 
- "ricker" is the Ricker-logistic model, `N[i,t] = N[i,t-1]*exp(gamma*(1-N[i,t-1]/omega))`, where gamma is the maximum instantaneous population growth rate (normally referred to as $r$) and omega is the equilibrium abundance (normally referred to as $K$). 
- "gompertz" is a modified version of the Gompertz-logistic model, `N[i,t] = N[i,t-1]*exp(gamma*(1-log(N[i,t-1]+1)/log(omega+1)))`, where the interpretations of gamma and omega are similar to in the Ricker model.

```{r}
# when using dynamics 'constant', gamma is constant
gamma <- 1 # recruitment rate

N <- Y <- matrix(NA, n, T)
S <- G <- matrix(NA, n, T - 1)
N[, 1] <- rpois(n, lambda) # 1st visit abundance
Y[, 1] <- rbinom(n, N[, 1], p) # 1st visit count
for (i in 1:(T - 1)) {
    S[, i] <- rbinom(n, N[, i], omega) # survival from t-1
    G[, i] <- rpois(n, gamma) # recruitment
    N[, i + 1] <- S[, i] + G[, i] # new N is survived + recruitment
    Y[, i + 1] <- rbinom(n, N[, i + 1], p) # new Y as observed
}
```

Estimate model parameters using `pcountOpen()`:

```{r}
## generalized N-mixture, T=3
visitMat <- matrix(as.character(1:3), n, 3, byrow = TRUE)
umfg <- unmarkedFramePCO(
    y = Y[, 1:3],
    siteCovs = x,
    obsCovs = list(visit = visitMat),
    numPrimary = 3
)
m0 <- pcountOpen(
    lambdaformula = ~ x1 + x2,
    gammaformula = ~1,
    omegaformula = ~1,
    pformula = ~ x3 + x4,
    data = umfg,
    K = K, mixture = "P",
    dynamics = "constant"
)
m0

cbind(
    est = coef(m0),
    true = c(beta, log(gamma), qlogis(omega), theta)
)
```

Single visit N-mixture model: we fit model to each primary period separately, ignoring the population dynamics

```{r}
## single visit with n obs, T=1
m1 <- svabu(Y[, 1] ~ x1 + x2 | x3 + x4, x,
    zeroinfl = FALSE, link.det = "logit", N.max = K
)
m2 <- svabu(Y[, 2] ~ x1 + x2 | x3 + x4, x,
    zeroinfl = FALSE, link.det = "logit", N.max = K
)
m3 <- svabu(Y[, 3] ~ x1 + x2 | x3 + x4, x,
    zeroinfl = FALSE, link.det = "logit", N.max = K
)
```


The assumption that transition parameters do not depend
on covariates is not realistic. It implies that as time passes, the
abundances are less affected by the covariates. However, when
these parameters do depend on covariates, the estimation procedure is extremely slow.

Populations are more open when omega is low, e.g. if survival is 0, all individuals come from recruitment (gamma).
On the other hand, when omega is 1 and gamma is 0, there is no turnover and the population is closed.
Omega dictates whether $N_{i} = N_{i,t}$ or not. Notice that the $N_{i} = N_{i,t}$
is the closed population assumption of the original N-mixture model.

When the population is open, visits become more independent, thus single visit
approach is suitable without forcing any population model onto the data.

We saw that detection probabilities varies over the visits.
We can fit a single visit model to all 3 visits at the same time,
thus providing more information for estimating the detection parameters, 
but we need to include visit as a factor in the abundance model to allow for 
variation in abundance across visits while keeping the detection estimates constant across visits.

```{r}
x123 <- rbind(
    data.frame(y = Y[, 1], visit = factor(1, 1:3), x),
    data.frame(y = Y[, 2], visit = factor(2, 1:3), x),
    data.frame(y = Y[, 3], visit = factor(3, 1:3), x)
)
m123 <- svabu(y ~ x1 + x2 + visit | x3 + x4, x123,
    zeroinfl = FALSE, link.det = "logit", N.max = K
)
```

Comparing the results:

```{r}
beta0 <- coef(m0)[grepl("lam\\(", names(coef(m0)))]
beta1 <- coef(m1, "sta")
beta2 <- coef(m2, "sta")
beta3 <- coef(m3, "sta")

theta0 <- coef(m0)[grepl("p\\(", names(coef(m0)))]
theta1 <- coef(m1, "det")
theta2 <- coef(m2, "det")
theta3 <- coef(m3, "det")

beta123 <- coef(m123, "sta")
theta123 <- coef(m123, "det")

lamt1 <- mean(lambda)
lamt2 <- mean(lamt1 * omega + gamma)
lamt3 <- mean(lamt2 * omega + gamma)
lam0 <- exp(drop(X %*% beta0))
lam01 <- mean(lam0)
lam02 <- mean(lam01 * plogis(coef(m0)["omega(Int)"])) +
    exp(coef(m0)["gamConst(Int)"])
lam03 <- mean(lam02 * plogis(coef(m0)["omega(Int)"])) +
    exp(coef(m0)["gamConst(Int)"])
lam1 <- mean(exp(drop(X %*% beta1)))
lam2 <- mean(exp(drop(X %*% beta2)))
lam3 <- mean(exp(drop(X %*% beta3)))
lam1v <- mean(exp(drop(cbind(X, 0, 0) %*% beta123)))
lam2v <- mean(exp(drop(cbind(X, 1, 0) %*% beta123)))
lam3v <- mean(exp(drop(cbind(X, 0, 1) %*% beta123)))

pt <- mean(p)
p0 <- mean(plogis(drop(Z %*% theta0)))
p1 <- mean(plogis(drop(Z %*% theta1)))
p2 <- mean(plogis(drop(Z %*% theta2)))
p3 <- mean(plogis(drop(Z %*% theta3)))
p123 <- mean(plogis(drop(Z %*% theta123)))

data.frame(
    truth = c(lam1 = lamt1, lam2 = lamt2, lam3 = lamt3, p1 = pt, p2 = pt, p3 = pt),
    DM = c(lam01, lam02, lam03, p0, p0, p0),
    SV = c(lam1, lam2, lam3, p1, p2, p3),
    SV123 = c(lam1v, lam2v, lam3v, p123, p123, p123)
)
```

When the population is fully open across primary periods and there is not
replication within primary periods, the generalized model has the same
identifiability requirements as the single visit model, it needs to satisfy the same RSPF condition.
