---
title: "Fitting N-mixture models"
subtitle: "Point count data analysis workshop 2025"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
# options(width = 53, scipen = 999)
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(unmarked)
    library(detect)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
})
set.seed(1234)
```

# Field data analysis

We will use the ABMI data, which has 3 equal (3.3 min) time intervals for each point count visit.
We use the point level data, but only for the center point, so we can keep 
sample size small and run the code quickly. When analyzing such data sets
for real, we would use all the data and not subset it like this.

```{r}
if (!interactive()) {
    d <- readRDS("../data/ab_birds_site_year_pt.rds")
} else {
    d <- readRDS("data/ab_birds_site_year_pt.rds")
}

spp <- "yellowrumped_warbler"

x <- data.frame(
    y1 = ifelse(d$visits$visit1[, spp] > 0, 1, 0),
    y2 = ifelse(d$visits$visit2[, spp] > 0, 1, 0),
    y3 = ifelse(d$visits$visit3[, spp] > 0, 1, 0),
    n1 = d$visits$visit1[, spp],
    n2 = d$visits$visit2[, spp],
    n3 = d$visits$visit3[, spp],
    d$vars
)
x$yday <- as.POSIXlt(x$fielddate)$yday / 365
x$hr <- x$fieldtime / 24
x$y_max <- pmax(x$y1, x$y2, x$y3)
x$n_max <- pmax(x$n1, x$n2, x$n3)
x <- x |>
    filter(
        hr > 0.1,
        hr < 0.6,
        yday > 0.37,
        yday < 0.5
    )
x0 <- x[x$point == 1, ]
```

# Single visit occupancy

```{r}
m_sv <- svocc(
    y1 ~ pconif + I(pconif^2) + thf | hr + I(hr^2) + yday,
    data = x0,
    link.sta = "logit"
)
summary(m_sv)

fit_sv2 <- plogis(
    model.matrix(formula(m_sv)$sta, x0) %*% coef(m_sv, model = "sta")
)
summary(fit_sv2)

data.frame(x0, p_occ = fit_sv2) |>
    ggplot(aes(x = pconif, y = p_occ, col = thf)) +
    geom_point() +
    ylim(0, 1)

d2 <- plogis(
    model.matrix(formula(m_sv)$det, x0) %*% coef(m_sv, model = "det")
)
summary(d2)

data.frame(x0, p_det = d2) |>
    ggplot(aes(x = hr, y = p_det, col = yday)) +
    geom_point() +
    ylim(0, 1)
data.frame(x0, p_det = d2) |>
    ggplot(aes(x = yday, y = p_det, col = hr)) +
    geom_point() +
    ylim(0, 1)

# sometimes it is hard to get SEs, use bootstrap
b <- bootstrap(m_sv, B = 250)
summary(b, type = "boot")
# explain parallelization options for bootstrap
```

Compare multiple and single visit results:

```{r}
umf2 <- unmarked::unmarkedFrameOccu(
    y = x0[, c("y1", "y2", "y3")],
    siteCovs = x0[, c("pconif", "thf")],
    obsCovs = list(
        hr = x0[, c("hr", "hr", "hr")],
        yday = x0[, c("yday", "yday", "yday")]
    )
)
m_mv2 <- unmarked::occu(
    ~ hr + I(hr^2) + yday ~ pconif + I(pconif^2) + thf,
    umf2
)
fit_mv2 <- plogis(model.matrix(m_mv2@formula[-2], x0) %*% coef(m_mv2, type = "state"))

data.frame(p_occ = fit_mv2, p_max = fit_sv2) |>
    ggplot(aes(x = fit_mv2, y = fit_sv2)) +
    geom_point() +
    geom_abline(lty = 2) +
    geom_smooth(method = "lm", se = FALSE)
mean(fit_mv2 / fit_sv2)
```

# Single visit N-mixture

```{r}
m_sv <- svabu(
    n1 ~ pconif + I(pconif^2) + thf | hr + yday,
    data = x0,
    N.max = 100,
    zeroinfl = FALSE
)
summary(m_sv)

fit_mv <- exp(
    model.matrix(formula(m_sv)$sta, x0) %*% coef(m_sv, model = "sta")
)
summary(fit_mv)
data.frame(x0, lam = fit_mv) |>
    ggplot(aes(x = pconif, y = lam, col = thf)) +
    geom_point()

d2 <- plogis(
    model.matrix(formula(m_sv)$det, x0) %*% coef(m_sv, model = "det")
)
summary(d2)
data.frame(x0, p_det = d2) |>
    ggplot(aes(x = hr, y = p_det, col = yday)) +
    geom_point()
data.frame(x0, p_det = d2) |>
    ggplot(aes(x = yday, y = p_det, col = hr)) +
    geom_point()

# try 2nd and 3rd visit, what if we stack visits and assume independence?

# calculate abundance from p_occ (using cloglog)

# show model selection
# explain strategy (why saturated to simpler works for SV)
# explain conditional likelihood for abundance and what it means for model selection (reduced param space)
```

# RSPF condition

RSPF condition reduces to the following two conditions:

1. not all covariates in the model are categorical, and 
2. the ratio function is nonlinear and involves all components of the
parameter vector

Remember that if the nonlinearity on the log-scale is weak, one
may need very large sample sizes to get reasonable estimates.

Use the shiny app to explore issues when both the status and detection components
of the model are constants:

```{r}
#| eval: false
shiny::runApp("apps/mvisit")
```

