---
title: "Single visit based approaches"
subtitle: "Analysing Ecological Data with Detection Error"
date: "`r Sys.Date()`"
author: "Péter Sólymos"
toc: true
format:
  html: 
    html-math-method: katex
    self-contained: true
  pdf: default
---

```{r}
#| include: false
# options(width = 53, scipen = 999)
library(knitr)
```

# Preamble

```{r}
suppressPackageStartupMessages({
    library(unmarked)
    library(detect)
    library(dplyr)
    library(ggplot2)
    theme_set(theme_light())
})
set.seed(1234)
```

# Field data analysis

We will use the ABMI data, which has 3 equal (3.3 min) time intervals for each point count visit.
We use the point level data, but only for the center point, so we can keep 
sample size small and run the code quickly. When analyzing such data sets
for real, we would use all the data and not subset it like this.

```{r}
if (!interactive()) {
    d <- readRDS("../data/ab_birds_site_year_pt.rds")
} else {
    d <- readRDS("data/ab_birds_site_year_pt.rds")
}

spp <- "yellowrumped_warbler"

x <- data.frame(
    y1 = ifelse(d$visits$visit1[, spp] > 0, 1, 0),
    y2 = ifelse(d$visits$visit2[, spp] > 0, 1, 0),
    y3 = ifelse(d$visits$visit3[, spp] > 0, 1, 0),
    n1 = d$visits$visit1[, spp],
    n2 = d$visits$visit2[, spp],
    n3 = d$visits$visit3[, spp],
    d$vars
)
x$yday <- as.POSIXlt(x$fielddate)$yday / 365
x$hr <- x$fieldtime / 24
x$y_max <- pmax(x$y1, x$y2, x$y3)
x$n_max <- pmax(x$n1, x$n2, x$n3)
x <- x |>
    filter(
        hr > 0.1,
        hr < 0.6,
        yday > 0.37,
        yday < 0.5
    )
x0 <- x[x$point == 1, ]
```

# Single visit occupancy

```{r}
m_sv <- svocc(
    y1 ~ pconif + I(pconif^2) + thf | hr + I(hr^2) + yday,
    data = x0,
    link.sta = "logit"
)
summary(m_sv)

fit_sv2 <- plogis(
    model.matrix(formula(m_sv)$sta, x0) %*% coef(m_sv, model = "sta")
)
summary(fit_sv2)

data.frame(x0, p_occ = fit_sv2) |>
    ggplot(aes(x = pconif, y = p_occ, col = thf)) +
    geom_point() +
    ylim(0, 1)

d2 <- plogis(
    model.matrix(formula(m_sv)$det, x0) %*% coef(m_sv, model = "det")
)
summary(d2)

data.frame(x0, p_det = d2) |>
    ggplot(aes(x = hr, y = p_det, col = yday)) +
    geom_point() +
    ylim(0, 1)
data.frame(x0, p_det = d2) |>
    ggplot(aes(x = yday, y = p_det, col = hr)) +
    geom_point() +
    ylim(0, 1)

# sometimes it is hard to get SEs, use bootstrap
b <- bootstrap(m_sv, B = 250)
summary(b, type = "boot")
# explain parallelization options for bootstrap
```

Compare multiple and single visit results:

```{r}
umf2 <- unmarked::unmarkedFrameOccu(
    y = x0[, c("y1", "y2", "y3")],
    siteCovs = x0[, c("pconif", "thf")],
    obsCovs = list(
        hr = x0[, c("hr", "hr", "hr")],
        yday = x0[, c("yday", "yday", "yday")]
    )
)
m_mv2 <- unmarked::occu(
    ~ hr + I(hr^2) + yday ~ pconif + I(pconif^2) + thf,
    umf2
)
fit_mv2 <- plogis(model.matrix(m_mv2@formula[-2], x0) %*% coef(m_mv2, type = "state"))

data.frame(p_occ = fit_mv2, p_max = fit_sv2) |>
    ggplot(aes(x = fit_mv2, y = fit_sv2)) +
    geom_point() +
    geom_abline(lty = 2) +
    geom_smooth(method = "lm", se = FALSE)
mean(fit_mv2 / fit_sv2)
```

# Single visit N-mixture

```{r}
m_sv <- svabu(
    n1 ~ pconif + I(pconif^2) + thf | hr + yday,
    data = x0,
    N.max = 100,
    zeroinfl = FALSE
)
summary(m_sv)

fit_mv <- exp(
    model.matrix(formula(m_sv)$sta, x0) %*% coef(m_sv, model = "sta")
)
summary(fit_mv)
data.frame(x0, lam = fit_mv) |>
    ggplot(aes(x = pconif, y = lam, col = thf)) +
    geom_point()

d2 <- plogis(
    model.matrix(formula(m_sv)$det, x0) %*% coef(m_sv, model = "det")
)
summary(d2)
data.frame(x0, p_det = d2) |>
    ggplot(aes(x = hr, y = p_det, col = yday)) +
    geom_point()
data.frame(x0, p_det = d2) |>
    ggplot(aes(x = yday, y = p_det, col = hr)) +
    geom_point()
```


_Exercise_: Try using `svabu()` with the 2nd and 3rd visits.

What if we stack visits and assume independence?

```{r}
#| eval: false
x1 <- x0 |>
    dplyr::select(n1, n2, n3, pconif, thf, hr, yday) |>
    tidyr::pivot_longer(
        cols = starts_with("n"),
        names_to = "visit",
        values_to = "n"
    )
m_sv123 <- svabu(
    n ~ pconif + I(pconif^2) + thf | hr + yday,
    data = x1,
    N.max = 100,
    zeroinfl = FALSE
)
summary(m_sv123)
```

Model selection is more complicated for single visit models, because we need to
make sure the model form satisfies the RSPF condition.

The following strategy is recommended for single visit models:

1. Fit the most complex model (saturated) for both the status and detection components
2. Then simplify the model by removing detection covariates one by one, starting with the least significant ones, and checking the AIC at each step
3. Make sure that detection probability has sufficient variation (e.g. by plotting it against covariates), i.e. it is not close to constant, otherwise the model may not be identifiable and estimates may be unreliable
4. Once you have a reasonable detection model, you can simplify the status component by removing covariates one by one, starting with the least significant ones, and checking the AIC at each step

```{r}
#| eval: false
m_sv_full <- svabu(
    n1 ~ pconif + I(pconif^2) + thf | hr + I(hr^2) + yday,
    data = x0,
    N.max = 100,
    zeroinfl = FALSE
)
summary(m_sv_full)

drop1(m_sv_full, model = "det") # no huge drop in AIC
m_sv_red <- update(m_sv_full, formula = . ~ . | . - I(hr^2))
summary(m_sv_red)
drop1(m_sv_red, model = "sta") # no better model
```

_Exercise_: Try `zeroinfl = TRUE` and `distr = "NB"` to explore other model forms.
Compare the models using AIC.

```{r}
#| eval: false
m_sv_P <- m_sv_red
m_sv_ZIP <- update(m_sv_P, zeroinfl = TRUE)
m_sv_NB <- update(m_sv_P, distr = "NB")
m_sv_ZINB <- update(m_sv_P, distr = "NB", zeroinfl = TRUE)

AIC(m_sv_P, m_sv_NB, m_sv_ZIP, m_sv_ZINB)
summary(m_sv_ZIP)
```

# RSPF condition

RSPF condition reduces to the following two conditions:

1. not all covariates in the model are categorical, and 
2. the ratio function is nonlinear and involves all components of the
parameter vector

Remember that if the nonlinearity on the log-scale is weak, one
may need very large sample sizes to get reasonable estimates.

Use the shiny app to explore issues when both the status and detection components
of the model are constants:

```{r}
#| eval: false
shiny::runApp("apps/mvisit")
```

